---
title: "Manitoba Assessment Parcels — 3D Value Map"
author: "Jason M. Schellenberg, P. App, AACI"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    toc: true
output-file: index.html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

Interactive 3D map of Manitoba's property assessment parcels.
**Height** encodes total assessed value; **color** encodes municipality.
Use the **search bar** (top left) to find any address in Manitoba.
Hover over a parcel to see its assessed value, roll number, and how it compares to the municipality median.
Use the **layer toggle** (bottom left) to switch to a municipality-level view comparing total assessed value across municipalities.

```{r}
#| label: setup
#| include: false

library(duckdb)
library(DBI)
library(freestiler)
library(mapgl)
library(arrow)
library(geoarrow)
library(dplyr)
library(sf)
library(fs)
library(glue)
library(cli)
library(scales)
library(readr)
library(stringr)

layer_name <- "mb-parcels"
pmtiles_file <- "data/mb-assessment-parcels.pmtiles"

# Set to TRUE to use local data embedded in the HTML (works offline, larger file).
# Set to FALSE to use PMTiles from GitHub Releases (smaller file, requires upload).
use_local_data <- TRUE

# PMTiles URL for the deployed map (GitHub Releases — supports HTTP range requests)
pmtiles_url <- "mb-assessment-parcels.pmtiles"  # Relative path — served from docs/ alongside index.html

# Assessment roll metadata
base_value_date <- "April 1, 2023"
```

```{r}
#| label: find-data
#| include: false

parquet_files <- fs::dir_ls("data", glob = "*/mb-assessment-parcels-*.parquet")

if (length(parquet_files) == 0) {
  stop("No GeoParquet files found in data/. Run MBAssessmentParcels.qmd first.")
}

input_file <- sort(parquet_files, decreasing = TRUE)[1]
file_date <- stringr::str_extract(basename(input_file), "\\d{4}-\\d{2}-\\d{2}")

# Extract assessment year from the Asmt_Roll field (e.g. "2026 Tax Assessment Roll")
asmt_year <- read_parquet(input_file, col_select = "Asmt_Roll") |>
  pull(Asmt_Roll) |>
  na.omit() |>
  head(1) |>
  stringr::str_extract("\\d{4}")

cli_alert_success("Using: {basename(input_file)}")
cli_alert_info("Download date: {file_date}")
cli_alert_info("File size: {fs::file_size(input_file)}")
```

## Provincial Overview — `r asmt_year` Assessment Roll (Base Value Date: `r base_value_date`)

```{r}
#| label: provincial-overview
#| include: false

parcels_all <- read_parquet(input_file) |> as.data.frame()

n_parcels <- nrow(parcels_all)
n_munis <- n_distinct(parcels_all$Muni_Name_With_Typ)
total_assessed <- sum(parcels_all$Total_Value_Num, na.rm = TRUE)
median_assessed <- median(parcels_all$Total_Value_Num, na.rm = TRUE)

# Top municipality by total assessed value
top_muni <- parcels_all |>
  group_by(Muni_Name_With_Typ) |>
  summarise(total = sum(Total_Value_Num, na.rm = TRUE), .groups = "drop") |>
  slice_max(total, n = 1)
```

Manitoba's provincial assessment roll covers **`r format(n_parcels, big.mark = ",")`** property parcels across **`r n_munis`** municipalities, with a combined assessed value of **`r dollar(total_assessed, accuracy = 1, scale = 1e-9, suffix = "B")`**. The median property assessment is **`r dollar(median_assessed, accuracy = 1)`**. The largest municipality by total assessed value is **`r top_muni$Muni_Name_With_Typ`** at `r dollar(top_muni$total, accuracy = 1, scale = 1e-9, suffix = "B")`.

*Note: This dataset covers Manitoba outside Winnipeg. Winnipeg maintains a [separate assessment roll](https://data.winnipeg.ca/Assessment-Taxation-Corporate/Assessment-Parcels/d4mq-wa44).*

Data source: [Manitoba Assessment Parcels](https://geoportal.gov.mb.ca/datasets/8106acf39b124422a5f03a5c4e55d269_0/explore) via the Manitoba GeoPortal (downloaded `r file_date`).

```{r}
#| label: load-map-data
#| include: false

cli_alert_info("Loading parcel geometry from {basename(input_file)}...")

parcels_sf <- read_parquet(input_file) |>
  mutate(geometry = st_as_sfc(geometry)) |>
  st_as_sf()

# Simplify geometry for faster rendering and smaller HTML output
# Simplify geometry for embedded HTML to reduce file size
# dTolerance = 0.0005 (~55m) keeps parcels recognizable while reducing vertex count
if (use_local_data) {
  sf_use_s2(FALSE)
  parcels_sf <- st_make_valid(parcels_sf) |>
    st_simplify(dTolerance = 0.0005, preserveTopology = TRUE)
  sf_use_s2(TRUE)
  cli_alert_info("Simplified geometry for local embedding")
}

# Compute municipality medians and join back for tooltip context
muni_medians <- parcels_sf |>
  st_drop_geometry() |>
  group_by(Muni_Name_With_Typ) |>
  summarise(Muni_Median = median(Total_Value_Num, na.rm = TRUE), .groups = "drop")

parcels_sf <- parcels_sf |>
  left_join(muni_medians, by = "Muni_Name_With_Typ") |>
  mutate(Muni_Median_Fmt = if_else(
    is.na(Muni_Median), "N/A", dollar(Muni_Median, accuracy = 1)
  )) |>
  select(Roll_No_Txt, Total_Value, Total_Value_Num,
         Muni_Median_Fmt,
         Property_Address, Muni_Name_With_Typ,
         Dwelling_Units)

cli_alert_success("Loaded {format(nrow(parcels_sf), big.mark = ',')} parcels")

# Dissolve parcels into municipality boundaries for choropleth layer
cli_alert_info("Dissolving parcels by municipality...")
sf_use_s2(FALSE)
muni_sf <- parcels_sf |>
  group_by(Muni_Name_With_Typ) |>
  summarise(
    Total_Assessed = sum(Total_Value_Num, na.rm = TRUE),
    Parcel_Count = n(),
    Median_Value = median(Total_Value_Num, na.rm = TRUE),
    .groups = "drop"
  ) |>
  st_make_valid() |>
  st_simplify(dTolerance = 0.001, preserveTopology = TRUE) |>
  st_make_valid() |>
  mutate(
    Total_Assessed_Fmt = dollar(Total_Assessed, accuracy = 1),
    Median_Value_Fmt = dollar(Median_Value, accuracy = 1),
    Value_Rank = paste(rank(-Total_Assessed, ties.method = "min"), "of", n())
  )
sf_use_s2(TRUE)
cli_alert_success("Created {nrow(muni_sf)} municipality polygons")
```

```{r}
#| label: generate-pmtiles
#| include: false
#| eval: !expr '!use_local_data'

# Skip if PMTiles is newer than source parquet
needs_rebuild <- !fs::file_exists(pmtiles_file) ||
  fs::file_info(pmtiles_file)$modification_time <
  fs::file_info(input_file)$modification_time

if (needs_rebuild) {
  cli_alert_info("Generating PMTiles from {basename(input_file)}...")

  freestile(
    input = parcels_sf,
    output = pmtiles_file,
    layer_name = layer_name,
    min_zoom = 5L,
    max_zoom = 14L
  )

  cli_alert_success("PMTiles written: {fs::file_size(pmtiles_file)}")
} else {
  cli_alert_info("PMTiles up to date: {fs::file_size(pmtiles_file)}")
}
```

```{r}
#| label: municipality-codes
#| include: false

parcels_df <- parcels_all |>
  count(Muni_Name_With_Typ, sort = TRUE)

cli_alert_info("Found {nrow(parcels_df)} municipalities")
parcels_df |> head(20) |> knitr::kable()
```

```{r}
#| label: define-palette
#| include: false

# Define a categorical color palette for the top municipalities
# Colors chosen for visual distinction on a dark basemap
palette <- c(
  "#e74c3c",   # red
  "#3498db",   # blue
  "#2ecc71",   # green
  "#f39c12",   # orange
  "#9b59b6",   # purple
  "#1abc9c",   # teal
  "#e67e22",   # dark orange
  "#ec407a",   # pink
  "#26c6da",   # cyan
  "#ffca28",   # amber
  "#8d6e63",   # brown
  "#78909c"    # blue-grey
)

# Use top municipalities by parcel count for color coding
muni_names <- parcels_df$Muni_Name_With_Typ
muni_colors <- rep_len(palette, length(muni_names))
```

## Interactive Map of Municipalities & Parcels

```{r}
#| label: map-3d

# Quantile breaks for municipality choropleth color scale
total_vals <- muni_sf$Total_Assessed[!is.na(muni_sf$Total_Assessed)]
q_breaks <- as.numeric(quantile(total_vals, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1.0)))

m <- maplibre(
  style = carto_style("dark-matter"),
  center = c(-98.0, 55.0),
  zoom = 5,
  pitch = 30,
  bearing = 0,
  projection = "mercator"
) |>
  add_navigation_control(position = "top-right") |>
  add_geocoder_control(
    position = "top-left",
    placeholder = "Search for an address..."
  )

# Municipality choropleth layer (hidden by default)
m <- m |>
  add_fill_layer(
    id = "muni-total-value",
    source = muni_sf,
    fill_color = interpolate(
      column = "Total_Assessed",
      values = q_breaks,
      stops = c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
    ),
    fill_opacity = 0.8,
    visibility = "none",
    tooltip = concat(
      "<strong>", get_column("Muni_Name_With_Typ"), "</strong><br>",
      "Total Assessed Value: ", get_column("Total_Assessed_Fmt"), "<br>",
      "Rank: ", get_column("Value_Rank"), "<br>",
      "Parcels: ", get_column("Parcel_Count"), "<br>",
      "Median Value: ", get_column("Median_Value_Fmt")
    ),
    hover_options = list(
      fill_color = "#ffffff",
      fill_opacity = 1
    )
  )

# 3D parcel extrusion layer (visible by default)
if (use_local_data) {
  m <- m |>
    add_fill_extrusion_layer(
      id = "mb-parcels-3d",
      source = parcels_sf,
      fill_extrusion_height = interpolate(
        column = "Total_Value_Num",
        values = c(0, 100000, 500000, 2000000, 10000000),
        stops = c(0, 50, 200, 600, 2000)
      ),
      fill_extrusion_color = match_expr(
        column = "Muni_Name_With_Typ",
        values = muni_names,
        stops = muni_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("Property_Address"), "</strong><br>",
        "Assessed Value: ", get_column("Total_Value"), "<br>",
        "Municipality Median: ", get_column("Muni_Median_Fmt"), "<br>",
        "Municipality: ", get_column("Muni_Name_With_Typ"), "<br>",
        "Roll No: ", get_column("Roll_No_Txt"), "<br>",
        "Dwelling Units: ", get_column("Dwelling_Units")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
} else {
  m <- m |>
    add_pmtiles_source(
      id = "mb-parcels-source",
      url = pmtiles_url
    ) |>
    add_fill_extrusion_layer(
      id = "mb-parcels-3d",
      source = "mb-parcels-source",
      source_layer = layer_name,
      fill_extrusion_height = interpolate(
        column = "Total_Value_Num",
        values = c(0, 100000, 500000, 2000000, 10000000),
        stops = c(0, 50, 200, 600, 2000)
      ),
      fill_extrusion_color = match_expr(
        column = "Muni_Name_With_Typ",
        values = muni_names,
        stops = muni_colors,
        default = "#95a5a6"
      ),
      fill_extrusion_opacity = 0.8,
      tooltip = concat(
        "<strong>", get_column("Property_Address"), "</strong><br>",
        "Assessed Value: ", get_column("Total_Value"), "<br>",
        "Municipality Median: ", get_column("Muni_Median_Fmt"), "<br>",
        "Municipality: ", get_column("Muni_Name_With_Typ"), "<br>",
        "Roll No: ", get_column("Roll_No_Txt"), "<br>",
        "Dwelling Units: ", get_column("Dwelling_Units")
      ),
      hover_options = list(
        fill_extrusion_color = "#ffffff"
      )
    )
}

# Build legend with abbreviated breakpoint labels (e.g. $1.2B, $450M)
legend_colors <- c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
fmt_short <- function(x) {
  dplyr::case_when(
    x >= 1e9  ~ paste0("$", trimws(format(round(x / 1e9, 1), nsmall = 1)), "B"),
    x >= 1e6  ~ paste0("$", trimws(format(round(x / 1e6, 0))), "M"),
    x >= 1e3  ~ paste0("$", trimws(format(round(x / 1e3, 0))), "K"),
    TRUE      ~ paste0("$", trimws(format(round(x, 0))))
  )
}
legend_labels <- fmt_short(q_breaks)

legend_rows <- paste0(sapply(seq_along(legend_colors), function(i) {
  label <- if (i == 1) {
    paste0("&lt; ", legend_labels[i + 1])
  } else if (i == length(legend_colors)) {
    paste0("&ge; ", legend_labels[i])
  } else {
    paste0(legend_labels[i], " \u2013 ", legend_labels[i + 1])
  }
  paste0(
    '<div style="display:flex;align-items:center;margin-bottom:3px;">',
    '<span style="display:inline-block;width:18px;height:12px;background:', legend_colors[i],
    ';border-radius:2px;margin-right:6px;flex-shrink:0;"></span>',
    '<span>', label, '</span></div>'
  )
}), collapse = "")

# Combine layer toggle + legend into ONE add_control (mapgl only keeps one custom control)
combined_control_html <- paste0(
  '<div style="background:rgba(0,0,0,0.75);padding:10px 12px;border-radius:6px;color:#fff;font-family:sans-serif;font-size:13px;line-height:1.6;">',
  '<div style="font-weight:bold;margin-bottom:4px;">View</div>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="map-layer" value="3d" checked style="margin-right:5px;">3D Parcels',
  '</label>',
  '<label style="display:block;cursor:pointer;">',
  '<input type="radio" name="map-layer" value="muni" style="margin-right:5px;">Total Value by Municipality',
  '</label>',
  '<div id="muni-legend" style="display:none;margin-top:8px;padding-top:8px;border-top:1px solid rgba(255,255,255,0.3);font-size:12px;line-height:1.5;">',
  '<div style="font-weight:bold;margin-bottom:5px;">Total Assessed Value</div>',
  legend_rows,
  '</div>',
  '</div>'
)

m <- m |>
  add_control(combined_control_html, position = "bottom-left") |>
  htmlwidgets::onRender("
    function(el, x) {
      var poll = setInterval(function() {
        if (!el.map || !el.map.isStyleLoaded()) return;
        clearInterval(poll);
        var map = el.map;
        var legend = document.getElementById('muni-legend');
        var radios = el.querySelectorAll('input[name=\"map-layer\"]');
        radios.forEach(function(radio) {
          radio.addEventListener('change', function() {
            if (this.value === '3d') {
              map.setLayoutProperty('mb-parcels-3d', 'visibility', 'visible');
              map.setLayoutProperty('muni-total-value', 'visibility', 'none');
              if (legend) legend.style.display = 'none';
            } else {
              map.setLayoutProperty('mb-parcels-3d', 'visibility', 'none');
              map.setLayoutProperty('muni-total-value', 'visibility', 'visible');
              if (legend) legend.style.display = 'block';
            }
          });
        });
      }, 200);
    }
  ")

m
```

## Assessed Value by Municipality — `r asmt_year` Assessment Roll (Base Value Date: `r base_value_date`)

```{r}
#| label: summary-stats

summary_table <- parcels_all |>
  group_by(Muni_Name_With_Typ) |>
  summarise(
    count = n(),
    total_value = sum(Total_Value_Num, na.rm = TRUE),
    median_value = median(Total_Value_Num, na.rm = TRUE),
    mean_value = mean(Total_Value_Num, na.rm = TRUE),
    max_value = if (all(is.na(Total_Value_Num))) NA_real_
                else max(Total_Value_Num, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_value))

summary_display <- summary_table |>
  mutate(
    total_value = dollar(total_value, accuracy = 1),
    median_value = dollar(median_value, accuracy = 1),
    mean_value = dollar(mean_value, accuracy = 1),
    max_value = dollar(max_value, accuracy = 1)
  )

DT::datatable(
  summary_display,
  colnames = c("Municipality", "Parcels", "Total Assessed Value",
               "Median", "Mean", "Max"),
  options = list(pageLength = 25, order = list(list(2, "desc"))),
  rownames = FALSE
)
```
